---
title: 'Exploratory data analysis'
subtitle: "Group number: G1D5"
author: "Nam Anh Le"
format: 
  html:
    number-sections: false
    code-tools: true
    embed-resources: true
---

```{r, message=FALSE}
library(tidyverse)
library(table1)
library(randomForest)
theme_set(theme_bw())

abalone <- read_csv(
  "abalone/abalone.data", 
  col_names = c("Sex", "Length", "Diameter", "Height", "Whole_Weight", "Shucked_Weight", "Viscera_Weight", "Shell_Weight", "Rings"
  )
)

```

# Dataset

The data has 4177 observations (with no missing attribute values), with the attributes being `Sex`, `Length`, `Diameter`, `Height`, `Whole_Weight`, `Shucked_Weight`, `Viscera_Weight`, `Shell_Weight`, and the dependent variable being `Rings` which is also mentioned to be a way to predict the age of an abalone.

```{r}
ggplot(abalone, aes(x = Rings)) + geom_histogram(binwidth = 1, boundary = 0.5) + 
  labs(
    title = "Distribution of Rings",
    x = "Number of Rings",
    y = "Count"
  )
```

The distribution of `Rings` is roughly bell-shaped, but it is positively skewed, so it does not appear to follow a perfectly normal distribution. Most abalones are concentrated around 8 to 11 rings, with fewer observations at very low or very high ring counts. There is also a noticeable right tail.

# Data Statistics

```{r}

abalone <- abalone |>
  mutate(
    Sex = factor(Sex, levels = c("M", "F", "I"),
                 labels = c("Male", "Female", "Infant"))
  )

table1(
  ~ Length + Diameter + Height + Whole_Weight + Shucked_Weight +
    Viscera_Weight + Shell_Weight + Rings | Sex,
  data = abalone,
  overall = "Total"
)
```

# Initial Visualizations

```{r}
ggplot(abalone, aes(y = Sex, x = Rings)) +
  geom_boxplot()
```

The boxplot suggests that infant abalones generally have fewer rings than male and female abalones. There are several high outliers in each group. Since the three groups overlap substantially, `Sex` alone does not appear to be a strong predictor of `Rings`.

```{r}

abalone |>
  pivot_longer(
    cols = c(Diameter, Height, Length),
    names_to = "variable",
    values_to = "values"
  ) |>
  ggplot(aes(x = values, y = Rings)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_grid(Sex ~ variable, scales = "free_x") +
  labs(
    title = "Relationships Between Diameter, Height, Length and Rings",
    x = "Measurement (mm)",
    y = "Number of Rings"
  )


abalone |>
  mutate(
    Sex = factor(Sex, levels = c("Male", "Female", "Infant"))
  ) |>
  pivot_longer(
    cols = c(Whole_Weight, Shucked_Weight, Viscera_Weight, Shell_Weight),
    names_to = "variable",
    values_to = "values"
  ) |>
  mutate(
    variable = factor(
      variable,
      levels = c("Whole_Weight", "Shucked_Weight", "Viscera_Weight", "Shell_Weight"),
      labels = c("Whole weight", "Shucked weight", "Viscera weight", "Shell weight")
    )
  ) |>
  ggplot(aes(x = values, y = Rings)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_grid(Sex ~ variable, scales = "free_x") +
  labs(
    title = "Relationships Between Weight Variables and Rings",
    x = "Weight (g)",
    y = "Number of Rings"
  )

```

Overall, the EDA suggests that both size and weight measurements are useful predictors of `Rings`. There is a consistent positive relationship across male, female, and infant abalones. However, most `Height` values are clustered within a relatively small range, approximately 0.0 to 0.3, with a few outliers. As a result, the fitted line for `Height` may be more affected by extreme observations and should be interpreted more cautiously.


```{r}
abalone$Sex <- as.factor(abalone$Sex)

set.seed(123)

rf_model <- randomForest(
  Rings ~ .,
  data = abalone,
  ntree = 500,
  importance = TRUE
)

rf_model
```

```{r}
importance(rf_model)
varImpPlot(rf_model)

```

```{r}
plot(abalone$Rings, rf_model$predicted,
     xlab = "Actual Rings",
     ylab = "Predicted Rings",
     main = "Random Forest: Actual vs Predicted")
abline(0, 1, col = "red")
```



```{r}
k <- 10

folds <- sample(rep(1:k, length.out = nrow(abalone)))

rmse_vals <- c()
mae_vals <- c()
r2_vals <- c()

for (i in 1:k) {
  train_data <- abalone[folds != i, ]
  test_data  <- abalone[folds == i, ]
  
  rf_fit <- randomForest(
    Rings ~ .,
    data = train_data,
    ntree = 500,
    importance = TRUE
  )
  
  preds <- predict(rf_fit, newdata = test_data)
  
  rmse_vals[i] <- sqrt(mean((preds - test_data$Rings)^2))
  mae_vals[i]  <- mean(abs(preds - test_data$Rings))
  r2_vals[i]   <- cor(preds, test_data$Rings)^2
}

mean(rmse_vals)
mean(mae_vals)
mean(r2_vals)
```





